---
title: "simulation_functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulation_functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = TRUE,
  tidy.opts=list(arrow=TRUE,width.cutoff = 50)
)
```


We present examples of setting up a simulation function for various models and R packages. These are written in a basic way to serve as blueprints for further customization.

The general layout of a simulation function is: 

```{r}
simfun <- function(N) {
    # Generate a data set
    # Test the hypothesis
}
```

For all examples in this vignette, we use an alpha level of .01 and a desired power of .95.

# T-Test

```{r}
library(mlpwr)
```

We test for one sample whether the mean differs from 0. The true effect size is Cohen's d = .3. In this scenario, this means that the true mean difference is .3.

```{r}
simfun_ttest <- function(N) {
    # Generate a data set
    dat <- rnorm(n = N, mean = 0.3)
    # Test the hypothesis
    res <- t.test(dat)
    res$p.value < 0.01
}
```

Example Use

```{r}
# res <- find.design(simfun = simfun_ttest,
#     boundaries = c(100,300), power = .95)
```

# ANOVA

```{r}
library(mlpwr)
```

We test for a mean difference among `n.groups` groups. Each group consists of `n` participants.

```{r}
simfun_anova <- function(n, n.groups) {

    # Generate a data set
    groupmeans <- rnorm(n.groups, sd = 0.2)  # generate groupmeans using cohen's f=.2
    dat <- sapply(groupmeans, function(x) rnorm(n,
        mean = x, sd = 1))  # generate data
    dat <- dat |>
        as.data.frame() |>
        gather()  # format

    # Test the hypothesis
    res <- aov(value ~ key, data = dat)  # perform ANOVA
    summary(res)[[1]][1, 5] < 0.01  # extract significance
}

```

Example Use

```{r}
# res <- find.design(simfun = simfun_anova, 
#   costfun = function(n,n.groups) 5*n+20*n.groups,
#   boundaries = list(n = c(10, 150), n.groups = c(5, 30)),
#   power = .95)
```

# Generalized Linear Models

```{r}
library(mlpwr)
```

We consider generalized linear models using the `stats` pacakge and the `glm` function.

## Using a fitted model

We use an "original" data set and fit a generalized linear model assuming a poisson distributed criterion `counts`.

```{r}
dat.original <- data.frame(
  counts = c(18, 17, 15, 20,
    10, 20, 25, 13, 12),
  treatment = gl(3, 1, 9),
  outcome = gl(3, 3))
mod.original <- glm(counts ~ outcome + treatment, data = dat.original,
    family = poisson)
summary(mod.original)
```

We use the parameters from the original model in the `simfun`.

```{r}
simfun_glm1 <- function(N) {

    # generate data
    dat <- data.frame(outcome = gl(3, 1, ceiling(N/3)),
        treatment = gl(3, ceiling(N/3)))[1:N, ]
    a <- predict(mod.original, newdata = dat, type = "response")
    dat$counts <- rpois(N, a)

    # test hypothesis
    mod <- glm(counts ~ outcome + treatment, data = dat,
        family = poisson)
    summary(mod)$coefficients["treatment2", "Pr(>|z|)"] <
        0.01
}
```

Example Use

```{r}
# res <- find.design(simfun = simfun_glm1,
#     boundaries = c(20,100), power = .95)
```

## Specifying parameters manually

We use a logistic regression and generate the data using hand-specified parameters and the logistic function.

```{r}
logistic <- function(x) 1/(1 + exp(-x))
```

We test if the second predictor is significant.

```{r}
simfun_glm2 <- function(N) {

    # generate data
    dat <- data.frame(pred1 = rnorm(N), pred2 = rnorm(N))
    beta <- c(1.2, 0.8)  # parameter weights
    prob <- logistic(as.matrix(dat) %*% beta)  # get probability
    dat$criterion <- runif(N) < prob  # draw according to probability

    # test hypothesis
    mod <- glm(criterion ~ pred1 + pred2, data = dat,
        family = binomial)
    summary(mod)$coefficients["pred2", "Pr(>|z|)"] <
        0.01
}
```

Example Use

```{r}
# res <- find.design(simfun = simfun_glm2,
#     boundaries = c(90,200), power = .95)
```

# Item Response Theory Models

```{r}
library(mlpwr)
library(mirt)
```

We use the `mirt` package to show an example that applies an item response theory model.

See `?simdata` for additional options and examples to generate data with the `mirt` package

## Testing a Rasch against a 2PL model

We first generate data from a 2PL  model. Then we want to find that the 2PL model actually fits the data better than a simpler Rasch model. We use a likelihood ratio test for this purpose. 

```{r}
simfun_irt1 <- function(N) {

    # generate data
    dat <- simdata(a = c(1.04, 1.2, 1.19, 0.61, 1.31,
        0.83, 1.46, 1.27, 0.51, 0.81), d = c(0.06,
        -1.79, -1.15, 0.88, -0.2, -1.87, 1.23, -0.08,
        -0.71, 0.6), N = N, itemtype = "2PL")

    # test hypothesis
    mod <- mirt(dat)  # Fit 2PL Model
    constrained <- "F = 1-4
          CONSTRAIN = (1-4, a1)"
    mod_constrained <- mirt(dat, constrained)  # Fit 2PL with equal slopes

    res <- anova(mod_constrained, mod)  # perform model comparison
    res$p[2] < 0.01  # extract significance
}
```

Example Use

```{r}
# res <- find.design(simfun = simfun_irt1,
#     boundaries = c(100,500), power = .95,evaluations =500)
```

## Testing for DIF

We detect differential item functioning in one item. Again, we generate data using the 2PL model, but using different parameters for two different groups in this case. The parameters for the first item are different in one group compared to other. We again apply an likelihood ratio test to test the null hypothesis that all item parameters are actually equal in both groups.

We optimize for the sizes of the two groups as the two design parameters. We assume that the second group involves higher costs than the first. 

```{r}
costfun_irt2 <- function(N1, N2) 5 * N1 + 7 * N2
```


```{r}
simfun_irt2 <- function(N1, N2) {

    # generate data
    a1 <- a2 <- c(1.04, 1.2, 1.19, 0.61, 1.31, 0.83,
        1.46, 1.27, 0.51, 0.81)
    d1 <- d2 <- c(0.06, -1.79, -1.15, 0.88, -0.2, -1.87,
        1.23, -0.08, -0.71, 0.6)
    a2[1] <- a2[1] + 0.3
    d2[1] <- d2[1] + 0.5
    dat1 <- simdata(a = a1, d = d1, N = N1, itemtype = "2PL")
    dat2 <- simdata(a = a2, d = d2, N = N2, itemtype = "2PL")
    dat <- as.data.frame(rbind(dat1, dat2))
    group <- c(rep("1", N1), rep("2", N2))

    # fit models
    mod1 <- multipleGroup(dat, 1, group = group)
    mod2 <- multipleGroup(dat, 1, group = group, invariance = c("slopes",
        "intercepts"))

    # test hypothesis
    res <- anova(mod2, mod1)

    # extract significance
    res$p[2] < 0.01
}
```

Example Use

```{r}
# res <- find.design(simfun = simfun_irt2,
#     boundaries = list(N1 = c(100,700), N2 = c(100,700)),
#     costfun = costfun_irt2,
#     power = .95)
```


# Multilevel Models

```{r}
library(mlpwr)
library(lme4)
library(lmerTest)
```

We consider multilevel models using the `lme4` and `lmerTest` pacakages. We use the `glmer` function for model fit and the `simulate` function to generate data. For further options for data generation in multilevel models see `?simulate.merMod`.

## Specifying parameters manually

We generate data using manually specified standard deviation of the random effects and parameter weights. We generate and fit the data according to a generalized linear mixed effects model with a poisson distributed criterion variable. 

```{r}
simfun_multi1 <- function(N) {

    # generate data
    params <- list(theta = 0.5, beta = c(2, -0.2, -0.4,
        -0.6))
    dat <- expand.grid(herd = 1:ceiling(N/4), period = factor(1:4))[1:N,
        ]
    dat$x <- simulate(~period + (1 | herd), newdata = dat,
        family = poisson, newparams = params)[[1]]

    # test hypothesis
    mod <- glmer(x ~ period + (1 | herd), data = dat,
        family = poisson)  # fit model
    pvalues <- summary(mod)[["coefficients"]][2:4,
        "Pr(>|z|)"]
    any(pvalues < 0.01)
}
```

Example Use

```{r}
# res <- find.design(simfun = simfun_multi1,
#     boundaries = c(100, 500),
#     power = .95)
```

## Using a fitted model 

We generate data from a fitted generalized linear mixed-effects model. We apply a mixed effects logistic regression in this case. It has two predictors and a random intercepts for each country.
 

```{r}
logistic <- function(x) 1/(1 + exp(-x))

N.original <- 300
n.countries.original <- 20

# generate original data
dat.original <- data.frame(country = rep(1:n.countries.original,
    length.out = N.original), pred1 = rnorm(N.original),
    pred2 = rnorm(N.original))
country.intercepts <- rnorm(n.countries.original, sd = 0.5)
dat.original$intercepts <- country.intercepts[dat.original$country]
beta <- c(1, 0.4, -0.3)  # parameter weights
prob <- logistic(as.matrix(dat.original[c("intercepts",
    "pred1", "pred2")]) %*% as.matrix(beta))  # get probability
dat.original$criterion <- runif(N.original) < prob  # draw according to probability

# fit original model to obtain parameters
mod.original <- glmer(criterion ~ pred1 + pred2 + 0 +
    (1 | country), data = dat.original, family = binomial)
```


In the simulation function, we generate criterion data using the original model. Design parameters are the number of participant per country `n` and the number of countries `n.countries`. We test the hypothesis that the second predictor is significant.

```{r}
simfun_multi2 <- function(n, n.countries) {

    # generate data
    dat <- data.frame(country = rep(1:n.countries,
        length.out = n * n.countries), pred1 = rnorm(n *
        n.countries), pred2 = rnorm(n * n.countries))
    dat$criterion <- simulate(mod.original, nsim = 1,
        newdata = dat, allow.new.levels = TRUE, use.u = FALSE) |>
        unlist()  # criterion data from the fitted model

    # test hypothesis
    mod <- glmer(criterion ~ pred1 + pred2 + 0 + (1 |
        country), data = dat, family = binomial)
    summary(mod)[["coefficients"]]["pred2", "Pr(>|z|)"] <
        0.01
}
```

As a cost function, we can use
```{r}
costfun_multi2 <- function(n, n.countries) 5 * m +
    100 * n.countries

```


Example Use

```{r}
# res <- find.design(simfun = simfun_multi2,
#     boundaries = list(n=c(10,40),n.countries=c(5,20)),
#     costfun = costfun_multi2,
#     power = .95)
```



